---
title: "**Analysis of GSS Categorical Data**"
author: "**Aryan Singh**"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(forcats)
```
Marital Status Analysis

Data Processing

Filter and clean marital status data to focus on three main categories:

```{r}
gss_cat %>% count(marital)
```


```{r}
marital_data <- gss_cat %>%
  select(marital) %>%
  filter(marital %in% c("Married", "Never Married", "Divorced")) %>%
  mutate(marital = fct_drop(marital))   # Remove unused factor levels
marital_table <- table(marital_data$marital)

```
Chi-Squared Goodness of Fit Test

```{r}
chisq.test(marital_table)
```

Visualization

```{r}
marital_data %>%
  ggplot(aes(x = marital)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Marital Status",
       x = "Marital Status",
       y = "Count") +
  theme_minimal()
```

Race Analysis

Data Processing

```{r}
race_tab <- gss_cat %>%
  select(race) %>%
  filter(race %in% c("White", "Black", "Other")) %>%
  mutate(race = fct_drop(race))

tab1 <- table(race_tab)  
```

Chi-Squared Test

```{r}
chisq.test(tab1)
```

Visualization

```{r}
ggplot(race_tab, aes(x = race)) +
  geom_bar(fill = "darkgreen") +
  labs(title = "Distribution of Race",
       x = "Race",
       y = "Count") +
  theme_minimal()
```

Party Identification Analysis

Data Processing

```{r}
party_tab <- gss_cat %>%
  select(partyid) %>%
  filter(partyid %in% c("Other party", "Strong republican", 
                       "Not str republican", "Ind,near rep",
                       "Independent", "Ind,near dem",
                       "Not str democrat", "Strong democrat")) %>%
  mutate(partyid = fct_drop(partyid))

tab2 <- table(party_tab)  
```

Chi-Squared Test

```{r}
chisq.test(tab2)
```

Visualization

```{r}
ggplot(party_tab, aes(x = partyid)) +
  geom_bar(fill = "red") +
  labs(title = "Distribution of Party Identification",
       x = "Party ID",
       y = "Count") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Religion Analysis

Data Processing

```{r}
religion_tab <- gss_cat %>%
  select(relig) %>%
  filter(relig %in% c("Inter-nondenominational", "Native american",
                     "Christian", "Orthodox-christian", "Moslem/islam",
                     "Other eastern", "Hinduism", "Buddhism", "Other",
                     "None", "Jewish", "Catholic", "Protestant")) %>%
  mutate(relig = fct_drop(relig))

tab3 <- table(religion_tab)  
```

Chi-Squared Test

```{r}
chisq.test(tab3)
```

Visualization

```{r}
ggplot(religion_tab, aes(x = relig)) +
  geom_bar(fill = "blue") +
  labs(title = "Distribution of Religious Affiliation",
       x = "Religion",
       y = "Count") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Denomination Analysis

Data Processing

```{r}
denomination_tab <- gss_cat %>%
  select(denom) %>%
  filter(denom %in% c("No denomination", "Other", "Episcopal", 
                     "Presbyterian-dk wh", "Presbyterian, merged", 
                     "Other presbyterian", "United pres ch in us", 
                     "Presbyterian c in us")) %>%
  mutate(denom = fct_drop(denom))

tab4 <- table(denomination_tab)  
```

Chi-Squared Test

```{r}
chisq.test(tab4)
```

Visualization

```{r}
ggplot(denomination_tab, aes(x = denom)) +
  geom_bar(fill = "purple") +
  labs(title = "Distribution of Denominations",
       x = "Denomination",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Conclusion

# Conclusion

All five categorical variables show highly non-uniform distributions.

- **Marital Status (df = 2):** χ² = 3359, p < .001 → Highly imbalanced categories
- **Race (df = 2):** χ² = 17956, p < .001 → Racial composition is dominated by one group
- **Party ID (df = 7):** χ² = 3772, p < .001 → Political affiliations are varied
- **Religion (df = 12):** χ² = 74133, p < .001 → Religious affiliation is extremely diverse
- **Denomination (df = 7):** χ² = 9469, p < .001 → Large differences in denominational identities


